---
title: "Week 8"
author: "Hao Ye"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
css: styles.css
output:
  html_document:
    number_sections: yes
    toc: yes
    smart: FALSE
---

```{r setup, include = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Objectives
* 3d plotting using `rgl`
    - basic plots with points and lines
    - setting plotting parameters
    - surface plots
    - output to file
    - R markdown hooks

# 3D plotting

The `rgl` package allows us to produce interactive 3-d plots. The basic commands are similar to the plotting functions in base R.

## Basic plots

The default plot3d uses points. Input is given as x, y, z coordinates as separate vectors, or as columns to a matrix or data.frame.

```{r}
library(rgl)
library(viridis)

n <- 1000
x <- sort(rnorm(n))
y <- rnorm(n)
z <- rnorm(n) + atan2(x, y)
plot3d(x, y, z, col = viridis(n))
```

Changing the plot type to `s` draws spheres instead of flat points.

```{r}
plot3d(x, y, z, type = "s", col = viridis(n), size = 1)
```

Changing the plot type to `l` draws connected lines.

```{r}
f <- function(x) {c(10 * (x[2] - x[1]), 
                    x[1] * (28 - x[3]) - x[2], 
                    x[1] * x[2] - 8/3 * x[3])}

points <- matrix(20, nrow = n, ncol = 3)
dt <- 0.01
for(t in 2:NROW(points))
{
    xx <- points[t-1,]
    k1 <- f(xx)
    k2 <- f(xx + dt/2 * k1)
    k3 <- f(xx + dt/2 * k2)
    k3 <- f(xx + dt/2 * k3)
    k4 <- f(xx + dt * k3)
    points[t,] <- xx + dt/6 * (k1 + 2*k2 + 2*k3 + k4)
}

dist_1 <- sqrt(rowSums((points - matrix(c(6*sqrt(2), 6*sqrt(2), 27), 
                                     nrow = n, ncol = 3, byrow = TRUE))^2))
dist_2 <- sqrt(rowSums((points - matrix(c(-6*sqrt(2), -6*sqrt(2), 27), 
                                     nrow = n, ncol = 3, byrow = TRUE))^2))
min_dist <- pmin(dist_1, dist_2)

plot3d(points, type = "l", col = hsv(h = (min_dist - min(min_dist)) / (max(min_dist) - min(min_dist)), s = 1, v = 1))
```

## Adding points and lines to a plot

Note that points and lines can be added to an existing plot using the `points3d` and `lines3d` functions respectively.

```{r}
x <- sort(rnorm(n))
y <- rnorm(n)
z <- rnorm(n)
plot3d(x, y, z, col = viridis(n), size = 2)

theta <- seq(0, 4*pi, length.out = n)
phi <- seq(0, 2*pi, length.out = n)
lines3d(cos(seq(0, 4*pi, length.out = n)), 
        sin(seq(0, 8*pi, length.out = n)), 
        cos(seq(0, 6*pi, length.out = n)), lwd = 5)

points3d(mean(x), mean(y), mean(z), size = 10)
```

## Plot parameters

To make plots reproducible, it's useful to be able to recreate specific plot settings, e.g. after resizing, rotating the interactive plot. `par3d` is the function that both sets and gets plot parameters. I found that the important ones are 

```{r}
plot3d(points, type = "l", col = hsv(h = (min_dist - min(min_dist)) / (max(min_dist) - min(min_dist)), s = 1, v = 1))

par3d()
```

```{r}
userMatrix <- matrix(c(0.8016672, 0.5976394, -0.01252565, 0, 
                       -0.1470179, 0.2174320, 0.96493983, 0, 
                       0.5794096, -0.7717192, 0.26217180, 0, 
                       0.0000000, 0.0000000, 0.00000000, 1), 
                     byrow = TRUE, nrow = 4)
zoom <- 1
windowRect <- c(2, 44, 944, 878)
scale <- c(1.0325553, 0.8813157, 1.1362212)
FOV <- 30
observer <- matrix(c(0, 0, 160), ncol = 3)
par3d(userMatrix = userMatrix, 
      zoom = zoom, 
      FOV = FOV, 
      scale = scale, 
      windowRect = windowRect)
observer3d(observer)
```

## Changing plot settings

Note that we can also control the display of axes and bounding boxes:

```{r}
plot3d(points, type = "l", col = hsv(h = (min_dist - min(min_dist)) / (max(min_dist) - min(min_dist)), s = 1, v = 1), 
       axes = FALSE, 
       xlab = "x", ylab = "", zlab = "z")
axes3d(c("x--", "y+-", "z--"))
mtext3d("y", "y+-", line = 2)
```

## Surface Plots

```{r}
data(volcano)

z <- 2 * volcano        # Exaggerate the relief
x <- 10 * (1:nrow(z))   # 10 meter spacing (S to N)
y <- 10 * (1:ncol(z))   # 10 meter spacing (E to W)

zlim <- range(y)
zlen <- zlim[2] - zlim[1] + 1

colorlut <- terrain.colors(zlen) # height color lookup table

col <- colorlut[ z - zlim[1] + 1 ] # assign colors to heights for each point

surface3d(x, y, z, color = col, back = "lines")
```

## Saving output to a file

An existing plot can be saved to a file using the `rgl.postscript` function, which can export several different formats

```{r}
plot3d(points, type = "l", col = hsv(h = (min_dist - min(min_dist)) / (max(min_dist) - min(min_dist)), s = 1, v = 1))

rgl.postscript("lorenz.pdf", fmt = "pdf")
```

## R markdown hooks

`rgl` plots are created in a separate window, which means that they don't normally show up in knitted output from R markdown. However, the `knitr` package contains hooks to embed `rgl` plots, either in static form (`hook_rgl`) or as an interactive webgl figure (`hook_webgl`).

```{r}
library(knitr)
knit_hooks$set(rgl = hook_rgl)
knit_hooks$set(webgl = hook_webgl)
```

```{r, rgl = TRUE}
plot3d(points, type = "l", col = hsv(h = (min_dist - min(min_dist)) / (max(min_dist) - min(min_dist)), s = 1, v = 1))
```

```{r, webgl = TRUE}
plot3d(points, type = "l", col = hsv(h = (min_dist - min(min_dist)) / (max(min_dist) - min(min_dist)), s = 1, v = 1))
```

# Resources

* 