---
title: "Week 5"
author: "Hao Ye"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
css: styles.css
output:
  html_document:
    number_sections: yes
    toc: yes
    smart: FALSE
---

```{r setup, include = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Objectives

* Explore new plot types through ggplot geoms
    + `geom_histogram`, `geom_density`, and `geom_violin`
    + `geom_bin2d`, `geom_hex` and `geom_density2d`
    + Error bars (`geom_errorbar` and `stat_summary`)
    + `geom_rect` and `geom_tile`
* Apply various plot tweaks
    + coordinate transformations (scaling, limits)
    + axis labels
* Make multi-panel plots using `facet_grid` (from `ggplot2`)
* Make multi-panel plots using `grid.arrange` (from `gridExtra`)

# Review of ggplot

## Load the dataset from week 3

```{r}
dat <- read.csv("../week-3/EconomistData.csv") # note the path
head(dat)
```

## Scatter plot with CPI on the x axis and HDI on the y axis

```{r}
library(ggplot2)

my_plot <- ggplot(dat, aes(x = CPI, y = HDI)) + 
    geom_point()
print(my_plot)
```

## Scatter plot, with colors and shapes for each region, and different theme

```{r}
my_plot <- ggplot(dat, aes(x = CPI, y = HDI, 
                           color = Region, shape = Region)) + 
    geom_point() + 
    scale_color_brewer(palette = "Dark2") + 
    theme_bw()
print(my_plot)
```

# More geom types

## 1-D summary of data

### Histograms

A simple histogram of `CPI` values using `geom_histogram`

```{r}
my_plot <- ggplot(dat, aes(x = CPI)) + 
    geom_histogram() + 
    theme_bw()
print(my_plot)
```

A stacked version, with different colors by region

```{r}
my_plot <- ggplot(dat, aes(x = CPI, fill = Region)) + 
    geom_histogram() + 
    scale_fill_brewer(palette = "Dark2") + 
    theme_bw()
print(my_plot)
```

### Density Map (1D)

If we want a smoothed histogram, where the density of the underlying distribution is inferred, we can use `geom_density` instead:

```{r}
my_plot <- ggplot(dat, aes(x = CPI, fill = Region)) + 
    geom_density(alpha = 0.5) + 
    scale_fill_hue() + 
    theme_bw()
print(my_plot)
```

### Violin Plots

Violin plots are similar to using `geom_density`, but in the fashion of `geom_boxplot`:

```{r}
my_plot <- ggplot(dat, aes(x = Region, y = CPI, color = Region)) + 
    geom_boxplot() + 
    scale_color_brewer(palette = "Dark2") + 
    theme_bw()
print(my_plot)
```

```{r}
my_plot <- ggplot(dat, aes(x = Region, y = CPI, color = Region)) + 
    geom_violin() + 
    scale_color_brewer(palette = "Dark2") + 
    theme_bw()
print(my_plot)
```

## 2-D summary of data

If we have two axes we want to visualize at the same time, then we need slightly different geoms.

### Histograms (2D)

```{r}
my_plot <- ggplot(diamonds, aes(x = carat, y = price)) + 
    geom_bin2d() + 
    theme_bw()
print(my_plot)
```

### Histogram (2D Hexes)

```{r}
my_plot <- ggplot(diamonds, aes(x = carat, y = price)) + 
    geom_hex() + 
    theme_bw()
print(my_plot)
```

### Density Map (2D)

```{r}
my_plot <- ggplot(diamonds, aes(x = carat, y = price)) + 
    geom_density2d(aes(color = ..level..)) + 
    theme_bw()
print(my_plot)
```

## Error Bars

We previously went over (very briefly), the use of `stat_bin` to display means and standard errors, but there are also specific geoms we can use to display error bars.

Here, we use jitter to space out the points, and then add in the mean and standard error as summary values:
```{r}
my_plot <- ggplot(dat, aes(x = Region, y = CPI, color = Region)) + 
    geom_point(size = 0.5, position = "jitter") + 
    stat_summary() + 
    scale_color_brewer(palette = "Dark2") + 
    theme_bw()
print(my_plot)
```

The point and line is a bit hard to see, so let's use an alternate geom:
```{r}
my_plot <- ggplot(dat, aes(x = Region, y = CPI, color = Region)) + 
    geom_point(size = 0.5, position = "jitter") + 
    stat_summary(geom = "crossbar") + 
    scale_color_brewer(palette = "Dark2") + 
    theme_bw()
print(my_plot)
```

A more typical plot might be just to plot the means as bars, with error bars. Note that we have to specify for `geom_bar` that we don't want to plot the raw values (since there are multiple points for each region), but rather to compute the mean for each group.

```{r}
my_plot <- ggplot(dat, aes(x = Region, y = CPI, fill = Region)) + 
    geom_bar(stat = "summary", fun.y = "mean") + 
    geom_errorbar(stat = "summary") + 
    scale_color_brewer(palette = "Dark2") + 
    theme_bw()
print(my_plot)
```

Note that the previous plot uses means and standard errors. If we wanted to plot confidence intervals, we need a slightly different function, in this case, from the `Hmisc` package.

```{r}
library(Hmisc)
my_plot <- ggplot(dat, aes(x = Region, y = CPI, fill = Region)) + 
    geom_bar(stat = "summary", fun.y = "mean") + 
    geom_errorbar(fun.data = mean_cl_normal, stat = "summary") + 
    scale_color_brewer(palette = "Dark2") + 
    theme_bw()
print(my_plot)
```

# Other Plot Tweaks

## Coordinate Transformations

If we want to zoom in (or out) on a portion of the plot, we can manually set the plot limits using `coord_cartesian`.

```{r}
my_plot <- ggplot(diamonds, aes(x = carat, y = price)) + 
    geom_hex() + 
    coord_cartesian(xlim = c(0, 4), ylim = c(0, 5000)) + 
    theme_bw()
print(my_plot)
```

Note that this is different from using `scale_x_continuous` (and similar), which actually restricts the data before plotting (which may cut off objects)

```{r}
my_plot <- ggplot(diamonds, aes(x = carat, y = price)) + 
    geom_hex() + 
    scale_x_continuous(limits = c(0, 4)) + 
    scale_y_continuous(limits = c(0, 5000)) + 
    theme_bw()
print(my_plot)
```

Similarly, we can set a log axis:

```{r}
my_plot <- ggplot(diamonds, aes(x = carat, y = price)) + 
    geom_hex() + 
    scale_y_log10() + 
    theme_bw()
print(my_plot)
```

# Multiple panels using facets

If we want to visualize separate plots for different categorical variables, we can use the facet system in ggplot:

```{r}
my_plot <- ggplot(dat, aes(x = HDI, y = CPI)) + 
    geom_point() + 
    facet_wrap(~ Region) + 
    theme_bw()
print(my_plot)
```

Note that by default, it uses the same axes, which facilitates visual comparison. But this can be turned off:

```{r}
my_plot <- ggplot(dat, aes(x = HDI, y = CPI)) + 
    geom_point() + 
    facet_wrap(~ Region, scales = "free") + 
    theme_bw()
print(my_plot)
```

# Multiple panels using different ggplot objects

Facets work when the plots are all the same type, but what if we want a generic multiple panel figure?

```{r}
library(gridExtra)

plot_1 <- ggplot(dat, aes(x = HDI, y = CPI)) + 
    geom_point() + 
    theme_bw()

plot_2 <- ggplot(dat, aes(x = HDI)) + 
    geom_density() + 
    theme_bw()

grid.arrange(plot_1, plot_2)
```

# Resources

* [ggplot gallery -- http://www.r-graph-gallery.com/portfolio/ggplot2-package/](http://www.r-graph-gallery.com/portfolio/ggplot2-package/)
* [extensions to ggplot (other packages) -- http://www.ggplot2-exts.org/gallery/](http://www.ggplot2-exts.org/gallery/)
* [various tweaks to prettify a ggplot figure -- http://zevross.com/blog/2014/08/04/beautiful-plotting-in-r-a-ggplot2-cheatsheet-3/](http://zevross.com/blog/2014/08/04/beautiful-plotting-in-r-a-ggplot2-cheatsheet-3/)