---
title: "Week 7"
author: "Hao Ye"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
css: styles.css
output:
  html_document:
    number_sections: yes
    toc: yes
    smart: FALSE
---

```{r setup, include = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Objectives
* Repetitive Tasks
    - `lapply` (base R) 
    - `map` (`purrr` package)
* R markdown chunk options
    - Evaluation: `eval`
    - Display output: `echo`, `include`, `results`
    - Saving output: `cache`
    - Figure options `fig`

# Repetitive Tasks

## `lapply` (base R)

`lapply` allows us to repeat a series of computations, specifically by making repeated function calls. The basic idea is that given a **vector** or **list** of inputs, `x`, and a function to apply for each input, `f`, produce the **list** of outputs, `y`.

In other words, each element, `y[i]` is computed as `f(x[i])`.

Because we use a function, each computation (generally) occurs independently of the others. This resolves some of the issues that can occur when using for-loops, by maintaining a consistent context for running the code in `f`.

```{r}
x <- 1:10
f <- sqrt # note that we want the actual function, so no ()
logp1 <- function(input) {log(input+1)}
y1 <- lapply(x, f)
y2 <- lapply(x, logp1)
```

Since the result is a list (safest form, since the outputs could be different types), if we want to combine them into a single data.frame using `unlist`.
```{r}
data.frame(x, unlist(y1), unlist(y2))
```

## `map` (`purrr` package)

In many cases, the `map` function from the `purrr` package operates the same as `lapply`:

```{r}
library(purrr)
z1 <- map(x, f)
z2 <- map(x, logp1)

data.frame(x, unlist(y1), unlist(y2), unlist(z1), unlist(z2))
```

Personally, I only use `purrr` for `map_df`, which combines outputs that are data frames into a single data frame. This is useful for e.g. doing complex computations that return summary data frames, on multiple inputs. (The alternative is a slightly clunky use of `do.call(rbind, )`).

```{r}
g <- function(input) {return(data.frame(mean = mean(input), uq = quantile(input, 0.75), lq = quantile(input, 0.25)))}

x <- list(rnorm(100), rnorm(1000), rnorm(10000))

map_df(x, g)
```

# Resources

