---
title: "Week 4"
author: "Hao Ye"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
css: styles.css
output:
  html_document:
    number_sections: yes
    toc: yes
    smart: FALSE
---

```{r setup, include = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Objectives

* Changing colors in ggplot
* The theme layer in ggplot
* Custom color palettes (`viridis`, `spaceMovie`)
* Adding statistical transformations and summaries to plots

# Review of ggplot

Previously, we introduced ggplot as a framework for constructing plots that emphasizes flexibility through its "grammar of graphics" system. Here we review some of the basic ideas.

## Load the dataset from last week

```{r}
dat <- read.csv("../week-3/EconomistData.csv") # note the path
head(dat)
```

## Scatter plot with CPI on the x axis and HDI on the y axis

```{r}
library(ggplot2)

my_plot <- ggplot(dat, aes(x = CPI, y = HDI))
my_plot + geom_point()
```

# Changing colors in ggplot

## Intro 

If we wanted to change the color of points in the plot, we might imagine two ways:
1. specify color in the geom layer (`geom_point` function)
2. specify color in the mapping layer (`aes` function)

What do you suppose each one does?

```{r, eval = FALSE}
my_plot <- ggplot(dat, aes(x = CPI, y = HDI, color = "blue"))
my_plot + geom_point()
```

```{r, eval = FALSE}
my_plot <- ggplot(dat, aes(x = CPI, y = HDI))
my_plot + geom_point(color = "blue")
```

## Basic Color Specification

Setting the color to be defined by the region variable produces a colored boxplot:
```{r}
my_plot <- ggplot(dat, aes(x = Region, y = HDI, color = Region))
my_plot + geom_boxplot()
```

But if we wanted to change the actual colors used, we need to specify what colors we want:
```{r}
my_plot <- ggplot(dat, aes(x = Region, y = HDI, color = Region))
my_plot + geom_boxplot() + scale_color_manual(values = c("red", "green", "blue", "magenta", "cyan", "yellow"))
```

And we can similarly define the exact mapping we want:
```{r}
my_plot <- ggplot(dat, aes(x = Region, y = HDI, color = Region))
my_plot + geom_boxplot() + 
    scale_color_manual(values = c("Asia Pacific" = "red", 
                                  "East EU Cemt Asia" = "green", 
                                  "MENA" = "blue", 
                                  "EU W. Europe" = "magenta", 
                                  "SSA" = "cyan", 
                                  "Americas" = "yellow"))
```

## What are the colors?

Colors in R can be found using the names listed in the output of "colors":
```{r}
head(colors())
```


And we can use these names when specifying colors:
```{r}
my_plot <- ggplot(dat, aes(x = Region, y = HDI))
my_plot + geom_boxplot(color = "midnightblue")
```

Alternatively, we can use one of the standard color spaces:
```{r}
my_color <- hsv(0.5, 0.75, 1.0) # <hue, saturation, value> color space
my_color <- rgb(0.25, 0.75, 0.5) # <red, green, blue> color space
my_color <- "#40C080" # using hexadecimal code
my_plot <- ggplot(dat, aes(x = Region, y = HDI))
my_plot + geom_boxplot(color = my_color)
```

# The theme layer in ggplot

In ggplot, the `theme` layer controls all the **non-data** aspects of a plot.

## Default themes

We can modify our plots by using some of the built-in themes:

```{r}
my_plot <- ggplot(dat, aes(x = Region, y = HDI, color = Region)) + geom_boxplot()
my_plot + theme_bw()
```

```{r}
my_plot + theme_dark()
```

## Custom themes

We can also customize the theme and specify changes to various components of the plot:

```{r}
my_plot + theme(
    axis.text = element_text(size = 14), 
    legend.key = element_rect(fill = "lightsalmon"), 
    legend.background = element_blank(), 
    panel.grid.minor = element_blank(), 
    panel.background = element_rect(fill = "lemonchiffon3")
)
```

Note that the modular nature of ggplot means you can save theme objects and re-use them, e.g. throughout a paper, or as part of a package, just like the default `theme_bw()`:

```{r}
my_theme <- theme(panel.grid.major = element_blank(),
                  panel.grid.minor = element_blank(),
                  axis.text = element_text(color = "black"), 
                  legend.background = element_rect(color = "black"), 
                  legend.key = element_blank(), 
                  legend.title = element_blank(), 
                  legend.position = c(0.5, 0), 
                  legend.justification = c(0.5, 0), 
                  legend.margin = margin(0.0, unit = "cm"), 
                  legend.key.size = unit(1.05, "lines"), 
                  legend.key.height = unit(1.05, "lines"), 
                  panel.background = element_rect(color = "black", fill = NA))

my_plot + my_theme
my_plot + my_theme + theme(legend.position = "none")
```

# Custom color palettes

Note that we previously specified a set of colors using `scale_color_manual` which is used to create a custom color scale (each color is manually specified.

## Built-in color palettes

As you might expect, ggplot has some built-in color options. Here are some of the ones to generate a **discrete** set of colors:

```{r}
my_plot + scale_color_grey()
```

```{r}
my_plot + scale_color_hue()
```

And some to generate a **continuous** set of colors:

```{r}
my_plot_2 <- ggplot(dat, aes(x = CPI, y = HDI, color = HDI)) + 
    geom_point()
my_plot_2 + scale_color_gradient2(low = "blue", mid = "black", high = "pink", midpoint = 0.6)
```

## Color palettes from packages

We can also use color palettes provided in packages:

```{r}
library(viridis)
my_plot + scale_color_viridis(discrete = TRUE)
my_plot_2 + scale_color_viridis(option = "A")
```

```{r}
library(spaceMovie)
my_plot + scale_color_manual(values = SW_palette("ANH"))
my_plot_2 + scale_color_gradientn(colors = SW_palette("AOTC"))
```

```{r}
my_plot + scale_color_brewer(palette = "Set3")
my_plot_2 + scale_color_distiller(palette = "RdYlGn")
```

## Exercises

1. Modify the HDI vs. CPI plot to use a different color and symbol for each region.

```{r}
ggplot(dat, aes(x = CPI, y = HDI, 
                color = Region, shape = Region)) + geom_point()
```

2. Modify the HDI boxplot to use a different color palette for the lines and for the fill.

```{r}
ggplot(dat, aes(x = Region, y = HDI, 
                color = Region, fill = Region)) + 
    geom_boxplot() + 
    scale_color_viridis(discrete = TRUE) + 
    scale_fill_viridis(discrete = TRUE, direction = -1)
```


# Adding statistic transformations and summaries to plots

By default, each `geom` has a `stat` component associated with it that describes how visually displays the plot. In `geom_point` for example, the `stat` is `identity`, because the x- and y- coordinates are being plotted as-is. In contrast, `geom_boxplot` must compute the bounds of the boxplot for the specified groups, and so that `stat` is `boxplot`.

For `geom_histogram`, the default `stat` is to count up the number of values in each bin:

```{r}
housing <- read.csv("../week-3/landdata-states.csv")
ggplot(housing, aes(x = Home.Value)) + 
    geom_histogram()
ggplot(housing, aes(x = Home.Value)) + 
    geom_histogram(stat = "bin", binwidth = 4000)
```

We can also add some of these statistical summaries to plots using `stat_*` functions instead of `geom_*` functions:

```{r}
my_plot + stat_summary_bin()
```

```{r}
my_plot_2 + stat_smooth()
```

## Exercises

1. Overlay a smoothing line on top of the scatter plot of HDI vs. CPI using `geom_smooth`, but use a linear model for the predictions. Hint: see `?stat_smooth`.

```{r}
ggplot(dat, aes(x = CPI, y = HDI)) + 
    geom_point() + 
    geom_smooth(method = "lm")
```

2. Overlay a smoothing line on top of the scatter plot using `geom_line`. Hint: change the statistical transformation.

```{r}
ggplot(dat, aes(x = CPI, y = HDI)) + 
    geom_point() + 
    geom_line(stat = "smooth", method = "loess")
```

3. Overlay a smoothing line on top of the scatter plot using the default `loess` method, but make it less smooth. Hint: see `?loess`.

```{r}
ggplot(dat, aes(x = CPI, y = HDI)) + 
    geom_point() + 
    geom_line(stat = "smooth", method = "loess", span = 0.2)
```

4. Try the above, but with different colors and smoothing for each `Region`.

```{r}
ggplot(dat, aes(x = CPI, y = HDI, 
                group = Region, color = Region)) + 
    geom_point() + 
    geom_smooth(method = "lm") + 
    scale_color_viridis(discrete = TRUE)
```

# Resources

* [theme layer documentation in ggplot -- http://docs.ggplot2.org/dev/vignettes/themes.html](http://docs.ggplot2.org/dev/vignettes/themes.html)
* [`spaceMovie` package for Star Wars palettes -- https://github.com/butterflyology/spaceMovie](https://github.com/butterflyology/spaceMovie)
* [`viridis` package for color palettes -- https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html](https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html)
* [Info about different color scales in ggplot -- http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/](http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/)
* [Cheatsheet about color palettes -- https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/colorPaletteCheatsheet.pdf](https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/colorPaletteCheatsheet.pdf)